Response to decision letter

### 2023-06-28 review.Rmd
add updated files for review submission

---
title: "Response to decision letter"
output: 
  pdf_document:
    keep_md: yes
date: "2023-05-01"
---

@@

**Please find below response to each point raised by the _Associate Editor_ Daniele Silvestro and the two _Reviewers_. We greatly appreciate your time and consideration for our resubmission.**

@@

**We added DOIs for Dryad package with Supplementary Material, L551.**

@@

**Done on L556-558.**

@@

**Thanks for this question. We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L221-228. Briefly, we implement the parsimony algorithm ACCTRAN using functions from the R package phangorn. We chose this algorithm as a quick way to obtain initial branch lengths that can be optimized afterwards using ML. The DNA matrices generated from data mined from the Barcode of Life Database (BOLD) do not have too much variation, so ACCTRAN resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson & Miller, 2008, doi:10.1111/j.1096-0031.2008.00229.x).Once parsimony branch lengths are estimated, they become parameters that are then optimized using Maximum Likelihood (ML), given the alignment, the topology and a simple JC model. The branches that are returned are ML branches.**
<!--ACCTRAN has been implemented in Forest et al. 2005. Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: a case study in the birch family Betulaceae. Systematic Botany.-->
**We added a vignette to the R package in which we showcase the whole workflow of branch length optimization, from parsimony to ML. This vignette is available at http://phylotastic.org/datelife/articles/making_bold_trees.html.**

@@

**Thanks for this questions. We elaborate on it on the "Description" section, at the end of subsection "Dating a tree topology with no branch lengths", L 202-210.Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths".**

**We want to point out that the main goal of the summarizing step in DateLife (when we distribute ages evenly in the chronogram) is to provide a single chronogram that can quickly show at a glance, and in the most agnostic way possible, the distribution of node ages obtained from published studies. Using an exponential distribution for missing node ages would make additional assumptions about the underlying evolutionary model of the chronogram, which can bias results in downstream analyses, as we elaborate in the discussion section, "Effects of phylogenetic sampling on downstream analyses" (L445 of manuscript without differences; L527 of manuscript with differences).**

@@

**Thanks for raising this point. We agree that this practice could be dangerous if a user is not aware that the age of the root was set in an arbitrary way. We changed the behaviour of the code to return a very conspicuous warning message when the root is arbitrary to make sure that users are aware of the fact (L196-201).**

@@

**Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L505) and Supplementary Material (L518-520).**

@@

**Thanks for pointing us to these references. We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of phylogenetic sampling on downstream analyses" (L481 of manuscript without differences; L527 of manuscript with differences), as well as changing the wording from "making up" to "in the absence of genetic data, simulating following a birth-death process" (L489; L540).**

@@

**We agree that users of all biological software should keep in mind that the results they get are only as good as the data they provide, and that they should implement some quality control. We mention this in the abstract and 3rd paragraph of discussion.**

@@

**Thanks for mentioning this. We think DateLife actually helps preventing this issue. As shown in Figure 5, DateLife allows to compare all age data available for a group, and users can immediately identify chronograms that are outliers, and explore whether it is related to an artifact of the chronogram or something else about the data or methodology used to generate that chronogram. Users can then choose to drop those calibrations from the final analysis.**

**We agree that it is a responsibility of the users to check the trees and chonograms before using them as secondary calibrations, in the same way researchers curate fossils to use as primary calibration points.**

@@

**The Associate Editor also pointed this out, and we agree that this feature of DateLife is not ideal, so we corrected it and address it in the code and in the manuscript. First, adding the 10% is too arbitrary. If there is more than one age data point, the function now adds one standard deviation unit of the mean of published ages to the maximum age value, and uses that as the root age. Second, the function now returns a conspicuous warning message when the root is not based on published data to make sure that users are aware of the fact (L196-201).**

@@

**Thanks for pointing that out. We added that topic to the intro along with a couple of references, L48-49.**

@@

**Thanks for pointing this out. We actually only ignore subspecies when retrieving data from a more inclusive taxonomic group. When provided by the user, subspecific taxa are processed and searched fully. We clarify this in the Description section "Creating a search query".**

@@

**It is indeed currently synced "manually". We specify this in the Description, L125. We comment in the discussion the benefits of having syncs done automatically, and ways to do so.**

@@

**We comment on this on L231-232. **

@@

**Indeed! The golden standard for obtaining divergence time estimates entails the use of primary calibrations (fossil or from geologic events).  Fossil preservation bias makes it impossible to get calibrations to constrain all nodes in a phylogenetic tree, hence, we will always have nodes that we need to infer using a model.**
**Moreover, uncertainty in placement of fossil calibrations and phylogenetic hypotheses makes it so that there are many different hypothesis supported by the golden standard that can be even conflicting.**
**The goal of DateLife is to provide the user with an insight on the current state of knowledge of time of lineage divergence for any group of taxa, as well as a tree summarizing the data.**

**We addressed that in the discussion.**

@@

**Thanks for pointing us to that paper! It gave us some nice ideas for discussion. We added a paragraph on the Discussion.**

@@

**Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi in addition to the GitHub addresses, L522-525.**

>>>>

---
title: "Response to decision letter"
output: 
  pdf_document:
    keep_md: yes
date: "2023-06-09"
---

@@

**Please find below response to each point raised by the _Associate Editor_ Daniele Silvestro and the two _Reviewers_. We experienced a delay due to a dependency to our software that was orphaned, and kicked out from the official R repository (CRAN), putting at risk the status of our own software. DateLife is hosted in other stable resources, but CRAN is the most widely used, so we decided to adopt the orphaned dependency to get it back on CRAN. However, the dependency is written in C, a language that we are not very familiar with, so it took us a while to figure it out. It was worthy, as our package is now stable in the official CRAN repository. We also carefully addressed all points made by the reviewers and AE, and we think they greatly improved the manuscript. We greatly appreciate your time and consideration for our resubmission.**

@@

**We added the DOI of our Dryad package containing all Supplementary Material on L690 (manuscript-differences.pdf).**

@@

**Done on L700 (manuscript-differences.pdf).**

@@

**Thanks for this question. We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L278 (manuscript-differences.pdf). Briefly, we implement the parsimony algorithm ACCTRAN using functions from the R package phangorn. We chose this algorithm as a quick way to obtain initial branch lengths that can be optimized afterwards using ML. The DNA matrices generated from data mined from the Barcode of Life Database (BOLD) do not have too much variation, so ACCTRAN resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson & Miller, 2008, doi:10.1111/j.1096-0031.2008.00229.x).Once parsimony branch lengths are estimated, they become parameters that are then optimized using Maximum Likelihood (ML), given the alignment, the topology and a simple JC model. The branches that are returned are ML branches.**
<!--ACCTRAN has been implemented in Forest et al. 2005. Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: a case study in the birch family Betulaceae. Systematic Botany.-->
**We added a vignette to the R package in which we showcase the whole workflow of branch length optimization, from parsimony to ML. This vignette is available at http://phylotastic.org/datelife/articles/making_bold_trees.html.**

@@

**Thanks for this questions. We elaborate on it on the "Description" section, at the end of subsection "Dating a tree topology without branch lengths", L 2268 (manuscript-differences.pdf). Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths".**

**We want to point out that the main goal of the summarizing step in DateLife (when we distribute ages evenly in the chronogram) is to provide a single chronogram that can quickly show at a glance, and in the most agnostic way possible, the distribution of node ages obtained from published studies. Using an exponential distribution for missing node ages would make additional assumptions about the underlying evolutionary model of the chronogram, which can bias results in downstream analyses, as we elaborate in the discussion section, "Effects of taxon sampling on downstream analyses" (L598 of file manuscript-differences.pdf).**

@@

**Thanks for raising this point. We agree that this practice could be dangerous if a user is not aware that the age of the root was set in an arbitrary way. We changed the behaviour of the code to return a very conspicuous warning message when the root is arbitrary to make sure that users are aware of the fact that an arbitrary age is assigned to the root (L262 manuscript-differences.pdf).**

@@

**Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L677 manuscript-differences.pdf) and Supplementary Material section (L5689 manuscript-differences.pdf).**

@@

**Thanks for pointing us to these references. We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of taxon sampling on downstream analyses" (L600 manuscript-differences.pdf), as well as changing the wording from "making up" to "completing a phylogeny using a stochastic birth-death polytomy".**

@@

**We agree that users of all biological software should keep in mind that the results they get are only as good as the data they provide, and that they should implement some quality control. We mention this in the abstract and 3rd paragraph of discussion L456 (manuscript-differences.pdf).**

@@

**Thanks for mentioning this. We think DateLife actually helps preventing this issue. As shown in Figure 5, DateLife allows to compare all age data available for a group, and users can immediately identify chronograms that are outliers, and explore whether it is related to an artifact of the chronogram or something else about the data or methodology used to generate that chronogram. Users can then choose to drop those calibrations from the final analysis. We mention this in several points of the disucssion.**

**Generaly, it is a responsibility of the users to check the trees and chonograms before using them as secondary calibrations, in the same way researchers curate fossils to use as primary calibration points.**

@@

**The Associate Editor also pointed this out, and we agree that this feature of DateLife is not ideal, so we corrected it and address it in the code and in the manuscript. First, adding the 10% is too arbitrary. If there is more than one age data point, the function now adds one standard deviation unit of the mean of published ages to the maximum age value, and uses that as the root age. Second, the function now returns a conspicuous warning message when the root is not based on published data to make sure that users are aware of the fact (L266 manuscript-differences.pdf).**

@@

**Thanks for pointing that out. We added that topic to the intro along with a couple of references, L71 manuscript-differences.pdf.**

@@

**Thanks for pointing this out. We actually only ignore subspecies when retrieving data from a more inclusive taxonomic group. When provided by the user, subspecific taxa are processed and searched fully. We clarify this in the Description section "Creating a search query" L104 manuscript-differences.pdf.**

@@

**It is indeed currently synced "manually". We specify this in the Description, L163 manuscript-differences.pdf. On the discussion, we elaborate on the benefits of having syncs done automatically, and ways to do so L446 manuscript-differences.pdf.**

@@

**We comment on this on L288 manuscript-differences.pdf. **

@@

**Indeed! We addressed that in the discussion L530 (manuscript-differences.pdf).**
**The golden standard for obtaining divergence time estimates entails the use of primary calibrations (fossil or from geologic events).  Fossil preservation bias makes it impossible to get calibrations to constrain all nodes in a phylogenetic tree, hence, we will always have nodes that we need to infer using a model.**
**Moreover, uncertainty in placement of fossil calibrations and phylogenetic hypotheses makes it so that there are many different hypothesis supported by the golden standard that can be even conflicting.**
**The goal of DateLife is to provide the user with an insight on the current state of knowledge of time of lineage divergence for any group of taxa, as well as a tree summarizing the data.**

@@

**Thanks for pointing us to that paper! It gave us some nice ideas for discussion. We added a paragraph on the Discussion L495 of manuscript-differences.pdf.**

@@

**Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi in addition to the GitHub URLs, L677 and L699 of manuscript-differences.pdf.**

### 2023-06-28 submission/final-captions.tex
add updated files for review submission

>>>>

Table 1

Ages of nodes extracted from 9 chronograms containing at least 2 of the 6 bird species used in the Fringillidae small example of the DateLife workflow. All nodes are congruent with the chosen tree topology. See Figure 3, step c2.

Table 2

Summary statistics of node ages from Table 1. Median ages are used by default to date a tree topology using DateLife, see Figure 3, step c3.
